Supplementary material: The increasing cost of happiness

### 2021-06-07 src/draft-SSM1.Rmd
Added validation of wellbeing measures. Closes issue #11 and #24

Affective wellbeing was determined by 9 questions in the SF-36 [(9a to 9i)](https://www.online.fbe.unimelb.edu.au/HILDAodd/KWCrossWaveCategoryDetails.aspx?varnt=gh9a). The SF-36 is a widely used self-completion measure of various aspects of physical, emotional and mental health [@ware2000sf]. A subset of 9 questions assess mental health and vitality, with four questions measuring agreement with positive aspects of mental health and vitality (i.e., "Feel full of life", "Felt calm and peaceful", "Have a lot of energy", "Been happy"), and five questions measuring agreement with negative aspects ("Felt so down in the dumps nothing could cheer me up", "Felt worn out", "Been a nervous person", "Felt down", "Felt tired"). The response scale timeframe was the past four weeks and agreement was indicated on a six-point Likert scale. Supplementary analysis on the subset of positive and negative items did not reveal any difference in the pattern of results between them (**insert Supplementary section name here**), and so we reverse scored negatively-phrased questions and calculated the sum of the nine questions so that higher scores represented better wellbeing. To aid interpretability, we rescaled the final sum to a score between 1-100, where 100 represents the maximum affective wellbeing achievable.

@@

We refer to the change point after which increases in income no longer produce similar increases in affective wellbeing as the _cost of happiness_. After this point, affective wellbeing is no longer as dependent on household income, and the economic security it represents. Presumably after this point further increases in affective wellbeing depend more on other life factors (e.g., leisure time, social connections) than financial security. Cognitive wellbeing on the other hand appeared to show consistent increases with household income and we found no evidence of any change point. The difference may reflect the importance of a numerical dollar value (e.g., bank balance, house value) when cognitively appraising one’s life achievements, versus the relevance of that number to our everyday experience of joy and our prevailing mood.

@@

>>>>

Affective wellbeing was determined by 9 questions in the SF-36 [(9a to 9i)](https://www.online.fbe.unimelb.edu.au/HILDAodd/KWCrossWaveCategoryDetails.aspx?varnt=gh9a). The SF-36 is a widely used self-completion measure of various aspects of physical, emotional and mental health [@ware2000sf], which has been validated in the Australian HILDA sample as a measure of health inequality [@butterworth2004validity]. A subset of 9 questions assess mental health and vitality, with four questions measuring agreement with positive aspects of mental health and vitality (i.e., "Feel full of life", "Felt calm and peaceful", "Have a lot of energy", "Been happy"), and five questions measuring agreement with negative aspects ("Felt so down in the dumps nothing could cheer me up", "Felt worn out", "Been a nervous person", "Felt down", "Felt tired"). The response scale timeframe was the past four weeks and agreement was indicated on a six-point Likert scale. We have previously shown a summed score of these 9 questions distinguishes the impact of good and bad major life events in a bidirectional manner [@kettlewell2020differential]. Supplementary analysis on the subset of positive and negative questions did not reveal any difference in the pattern of results between them (Supplementary materials, Figure S3), and so we reverse scored negatively-phrased questions and calculated the sum of the nine questions so that higher scores represented better wellbeing. To aid interpretability, we rescaled the final sum to a score between 1-100, where 100 represents the maximum affective wellbeing achievable.

@@

We refer to the change point after which increases in income no longer produce similar increases in affective wellbeing as the _cost of happiness_. After this point, affective wellbeing is no longer as dependent on household income, and the economic security it represents. Presumably after this point further increases in affective wellbeing depend more on other life factors (e.g., leisure time, social connections) than financial security. Cognitive wellbeing on the other hand appeared to show consistent increases with household income and we found no evidence of any change point. We have previously observed that cognitive- and affective-wellbeing appear to track distinct responses to financial gains and losses [@kettlewell2020differential], such that a major financial windfall produced less benefit to affective-wellbeing than cognitive-wellbeing whilst a major financial loss (e.g., bankruptcy) produced equal effects on both. The difference may reflect the importance of a numerical dollar value (e.g., bank balance, house value) when cognitively appraising one’s life achievements, versus the relevance of that number to our everyday experience of joy and our prevailing mood.

@@

#### 4.1 Limitations of the present study. 
The amount of variance in wellbeing explained by income was small. While the amount of variance was small, the pre-change slope was reliably above zero and small changes in happiness over an entire population will aggregate. Moreover, financial gains and losses have one of the largest impacts on subjective wellbeing in within-subject/fixed-effect models [e.g., @kettlewell2020differential], and there are many reasons why cross-sectional estimates of the covariance would under-represent this. Another limitation is that wellbeing is measured on a subjective scale, and we do not know whether small changes at different points on the scale represent larger changes in some other functional/real-world outcome (e.g., risk of suicide). At present, the real-world impact of changes in a subjective measure such as wellbeing is difficult to determine.

### 2021-06-07 src/supplementary-SSM1.Rmd
Added validation of wellbeing measures. Closes issue #11 and #24

bind_rows(parameter_summary_0c, parameter_summary_5c) %>%
  ggplot(aes(y = year, x = expected, color = model)) +
  geom_pointrange(aes(xmin = low, xmax = upp), position = position_dodge(width = 0.75)) +
  scale_y_continuous(breaks = c(2001, 2007, 2013, 2019)) +
  facet_wrap(~param, scales = "free_x") +
  ft_theme(base_size = 12) +
  scale_color_manual(values = c("#C79999", "grey65")) +
  theme(legend.position = "none",
        axis.title = element_blank())
```
```
Figure legend: Comparison of the conditional parameter estimates (red) and unconditional parameter estimates (grey) in the change point model of income on happiness. The conditional models included covariates for age, age squared, sex, education, and illness. Horizontal bar represents the 95% credible region and the solid point indicates the expected value (median) of each distribution.
```

@@

@@

##### Figure S3. Prior predictive check  
```{r figure_s3, fig.width=6, fig.height = 6, dpi=300}
load("../results/prior_predictions.Rdata")

@@

##### Figure S4. Residual plots for piecewise happiness models
```{r figure_s4, fig.height=9, fig.width=9, dpi=600}
# https://www.flutterbys.com.au/stats/tut/tut7.2b.html
# 
# 
# residual_fits <- map_dfr(nl_fits, ~{
#   
#   d0 <- resid(.x) %>%
#     as_tibble() %>%
#     select(resids = Estimate)
#   
#   d1 <- fitted(.x) %>%
#     as_tibble() %>%
#     select(fitteds = Estimate)
#   
#   bind_cols(d0, d1)
#   
#   }, .id = "fit") %>%
#   mutate(year = extract_numeric(fit))
# 
# save(residual_fits, file = paste0(here, "results/residual_fits.Rdata"))

@@

##### Figure S5. Happiness (SD units) at $50K/yr and $75K/yr from 2001-2019  
```{r figure_s5, fig.width=7, fig.height=5, dpi=300}
nl_fits = read_rds("../results/gh9_hhoecd_0cov_tight.RDS")

@@

model_summaries_all <- rbind(
  broom::glance(lm_fit) %>% mutate(model = "lin", year = "All"),
  broom::glance(lm_fit_c) %>% mutate(model = "linc", year = "All"),
  broom::glance(log_fit) %>% mutate(model = "log", year = "All"),
  broom::glance(log_fit_c) %>% mutate(model = "logc", year = "All"),
  broom::glance(pw_fit) %>% mutate(model = "pw", year = "All"),
  broom::glance(pw_fit_c) %>% mutate(model = "pwc", year = "All"),
  model_summaries
)
```

@@

left_join(lm_coef, log_coef, by = "Year") %>%
  left_join(pw_coefs, by = "Year") %>%
  mutate(Year = as.character(2000 + as.numeric(Year))) %>%
  flextable() %>%
  set_caption("OLS coefficients from linear, log-linear and piecewise models") %>%
  autofit()
```

>>>>

bind_rows(parameter_summary_0c, parameter_summary_5c) %>%
  ggplot(aes(y = year, x = expected, color = model)) +
  geom_pointrange(aes(xmin = low, xmax = upp), position = position_dodge(width = 0.75)) +
  scale_y_continuous(breaks = c(2001, 2007, 2013, 2019)) +
  facet_wrap(~param, scales = "free_x") +
  ft_theme(base_size = 12) +
  scale_color_manual(values = c("#C79999", "grey65")) +
  theme(legend.position = "none",
        axis.title = element_blank())
```
```
Figure legend: Comparison of the conditional parameter estimates (red) and unconditional parameter estimates (grey) in the piecewise-linear model of income on happiness. The conditional models included covariates for age, age squared, sex, education, and illness. Horizontal bar represents the 95% credible region and the solid point indicates the expected value (median) of each distribution.
```

@@

**Positive and negative components of affective wellbeing**. Affective wellbeing includes both positive and negative components, and both these components are assessed by multiple quesstions in the SF-36 item 9. To determine whether the increasing change-point in income on affective wellbeing we observed was due to positive or negative affect, we fit a piecewise-linear model of income on positive and negative affect separately. The four questions dealing with positive affect (i.e., "Feel full of life", "Felt calm and peaceful", "Have a lot of energy", "Been happy") and the five negative affect questions ("Felt so down in the dumps nothing could cheer me up", "Felt worn out", "Been a nervous person", "Felt down", "Felt tired") were summed together separately to create distinct scores for positive and negative affect. The negative affect score was reversed so that higher scoes indicated less negative affect. Household income was then regressed against each score using the piecewise linear model (Equation 3), as described for the overall affective wellbeing score in the main text. Figure S3 below shows that the change-point in income increased over the period 2001-2019 for both positive (pink) and negative (blue) affect, indicating a similar trend in both components of affective wellbeing over the 19 years.

##### Figure S3. Positive and negative affect parameters  
```{r figure_s3, fig.width=6, fig.height=8, dpi=300}
gh9neg.fits <- read_rds("../results/fits/excl_topcodes/sd/gh9neg_hhoecd_0cov_tight.RDS")
gh9pos.fits <- read_rds("../results/fits/excl_topcodes/sd/gh9pos_hhoecd_0cov_tight.RDS")

posteriors_pos <- map_dfr(gh9pos.fits, as.data.frame, .id = "fit") %>%
  left_join(income_params) %>%
  mutate(omega_dollars = (b_omega_Intercept * sd + mean)/1000) %>%
  as_tibble()

posteriors_pos %>%
  select(fit, contains("b0"), contains("b1"), contains("b2"), omega_dollars) %>%
  gather(key = "param", val = "sample", -fit) %>%
  mutate(
    year = extract_numeric(fit),
    param = recode_factor(param, !!!param_key)
  ) %>% 
  group_by(param, year) %>%
  summarise(
    low = quantile(sample, probs = 0.05, na.rm=T),
    expected = quantile(sample, probs = 0.5, na.rm=T),
    upp = quantile(sample, probs = 0.95, na.rm=T)
  ) %>%
  mutate(model = "positive") -> parameter_summary_pos

posteriors_neg <- map_dfr(gh9neg.fits, as.data.frame, .id = "fit") %>%
  left_join(income_params) %>%
  mutate(omega_dollars = (b_omega_Intercept * sd + mean)/1000) %>%
  as_tibble()

posteriors_neg %>%
  select(fit, contains("b0"), contains("b1"), contains("b2"), omega_dollars) %>%
  gather(key = "param", val = "sample", -fit) %>%
  mutate(
    year = extract_numeric(fit),
    param = recode_factor(param, !!!param_key)
  ) %>% 
  group_by(param, year) %>%
  summarise(
    low = quantile(sample, probs = 0.05, na.rm=T),
    expected = quantile(sample, probs = 0.5, na.rm=T),
    upp = quantile(sample, probs = 0.95, na.rm=T)
  ) %>%
  mutate(model = "negative") -> parameter_summary_neg

bind_rows(parameter_summary_pos, parameter_summary_neg) %>%
  ggplot(aes(y = year, x = expected, color = model)) +
  geom_pointrange(aes(xmin = low, xmax = upp), position = position_dodge(width = 0.75)) +
  scale_y_continuous(breaks = c(2001, 2007, 2013, 2019)) +
  facet_wrap(~param, scales = "free_x") +
  ft_theme(base_size = 12) +
  scale_color_manual(values = c("steelblue", "salmon")) +
  theme(legend.position = "none",
        axis.title = element_blank())
```
```
Figure legend: Comparison of the parameter estimates from positive items (pink) and negative items (blue) in the piecewise-linear model of income on affective wellbeing. The positive model regressed income against a sum of positive affect questions in the SF-36, while the negative model regressed income against a sum of negative affect questions (reversed scaled to match the positive affect score). Horizontal bar represents the 95% credible region and the solid point indicates the expected value (median) of each distribution.
```

Regressing income against the positive and negative components of affective wellbeing separately did not reveal substantially different trends for each. Both showed an increase in change-point income value over the 19 years, which by 2019 was credibly different from the base values in 2001. Changes in the other slope and intercept parameters were either not substantial or not credibly different from the 2001 values according to the 95% credible intervals.

@@

##### Figure S4. Prior predictive check  
```{r figure_s4, fig.width=6, fig.height = 6, dpi=300}
load("../results/prior_predictions.Rdata")

@@

##### Figure S5. Residual plots for piecewise happiness models
```{r figure_s5, fig.height=9, fig.width=9, dpi=600}
# https://www.flutterbys.com.au/stats/tut/tut7.2b.html
# 
# 
# residual_fits <- map_dfr(nl_fits, ~{
#   
#   d0 <- resid(.x) %>%
#     as_tibble() %>%
#     select(resids = Estimate)
#   
#   d1 <- fitted(.x) %>%
#     as_tibble() %>%
#     select(fitteds = Estimate)
#   
#   bind_cols(d0, d1)
#   
#   }, .id = "fit") %>%
#   mutate(year = extract_numeric(fit))
# 
# save(residual_fits, file = paste0(here, "results/residual_fits.Rdata"))

@@

##### Figure S6. Happiness (SD units) at $50K/yr and $75K/yr from 2001-2019  
```{r figure_s6, fig.width=7, fig.height=5, dpi=300}
nl_fits = read_rds("../results/gh9_hhoecd_0cov_tight.RDS")

@@

model_summaries_all <- rbind(
  broom::glance(lm_fit) %>% mutate(model = "lin", year = "Overall"),
  broom::glance(lm_fit_c) %>% mutate(model = "linc", year = "Overall"),
  broom::glance(log_fit) %>% mutate(model = "log", year = "Overall"),
  broom::glance(log_fit_c) %>% mutate(model = "logc", year = "Overall"),
  broom::glance(pw_fit) %>% mutate(model = "pw", year = "Overall"),
  broom::glance(pw_fit_c) %>% mutate(model = "pwc", year = "Overall"),
  model_summaries
)
```

@@

left_join(lm_coef, log_coef, by = "Year") %>%
  left_join(pw_coefs, by = "Year") %>%
  mutate(Year = as.character(2000 + as.numeric(Year))) %>%
  flextable() %>%
  fontsize(size = 9.5, part = "all") %>%
  set_caption("OLS coefficients from linear, log-linear and piecewise models") %>%
  autofit()
```
